---
output: 
  github_document:
    toc: false
---

[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/o2groups)](https://cran.r-project.org/package=o2groups)


# Overview

Effective outbreak control relies on a comprehensive understanding of how diseases spread between affected groups, such as healthcare workers and patients or specific age groups.

We developed an estimator that utilises information on transmission chains (who infected whom), enabling the identification and quantification of transmission patterns between groups.

The `o2groups` package provides a framework for simulating outbreaks using branching processes specifically designed for scenarios involving multiple groups with various transmission assortativity coefficients (`delta`).

The package is still under development and is not yet available on CRAN. 


```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

# Installation

```{r}
#devtools::install_github("CyGei/o2groups")
library(o2groups)
library(tidyverse)
```

# The assortativity coefficient
By quantifying the extent to which individuals transmit infections within their own groups, assortativity coefficients offer a means to measure the relative contributions of different groups to transmission. 
$\gamma$  is defined as the excess probability of a secondary infection taking place within group a compared to random expectations. $\gamma$  values range from 0 (fully disassortative) to ∞ (fully assortative), with 1 indicating homogeneous patterns. For instance, $\gamma$  = 2 indicates that an infected individual from group a is twice as likely to infect an individual from the same group compared to an individual from another group. Conversely, a $\gamma$  of 1/2  means that an infected individual from group a is twice as likely to infect an individual from another group compared to an individual from the same group. 

To simplify interpretation, we introduce a rescaled parameter $\delta$, ranging between -1 (fully disassortative) and 1 (fully assortative), with 0 corresponding to a homogeneous transmission pattern.

To convert $\gamma$ to $\delta$ (or vice versa), we use the following function:
  
```{r}
  gamma_values <- c(2, 1, 1/2)
  delta <- o2groups::scale(gamma_values)
  delta
  gamma <- o2groups::reverse_scale(delta)
  gamma

  # Plot the relationship between gamma and delta
  plot(seq(0, 100, 0.1), o2groups::scale(seq(0, 100, 0.1)), type = "l", xlab = "gamma", ylab = "delta")

```


# Example

## Simulation

We first specify the input parameters for our simulation.

```{r}
set.seed(123)
duration = 100 # the duration of the outbreak
n_groups = 3 # the number of groups
size = c(220, 200, 300) # the size of each group
name = c("HCW", "Inpatient", "Outpatient")
delta = o2groups::reverse_scale(c(0.5, 0.75, -0.5)) # the assortativity coefficient for each group
intro_n = c(1, 3, 1) # the number of introductions for each group
r0 = c(1.7, 1.5, 3) # the basic reproduction number for each group
generation_time = c(0, 0.1, 0.2, 0.4, 0.2, 0.1, 0) # the generation time distribution
incubation_period = round(rgamma(1000, shape = 2, scale = 1.5) # random values for the incubation period
) 
```

We then run the outbreak simulation using the `simulate_groups` function.

```{r}
set.seed(123)
sim <- simulate_groups(
  duration = duration,
  n_groups = n_groups,
  size = size,
  name = name,
  delta = delta,
  intro_n = intro_n,
  r0 = r0,
  generation_time = generation_time,
  incubation_period = incubation_period
)
```

## Visualisation of the outbreak 

We can first visualise the transmisson tree using the `plot_tree` function. 

```{r}

# Plot tree
o2groups::plot_tree(sim$data, pal = c("HCW" = "red3", "Inpatient" = "green4", "Outpatient" = "steelblue"))
```

As well as the incidence and proportion of susceptibles over time using the `plot_stats` function.

```{r}

# Plot incidence
o2groups::plot_stats(sim$stats)[1] # peak around day 14
# Plot proportion of susceptibles
o2groups::plot_stats(sim$stats)[2]

```

## Estimation of the assortativity coefficients 

We can estimate the assortativity coefficients using the `early_delta` function
```{r message=FALSE}

unscaled_est <- o2groups::early_delta(
  sim$data,
  min_t = 0,
  max_t = 13, #~ epidemic peak date
  name = name,
  size = size
)

unscaled_est # the estimated assortativity coefficients (gamma) and 95% CI for each group

```
We can plot the estimated assortativity coefficients and their 95% CI against the true values.
We analysed transmission chains up to the epidemic peak, which is around day 13. 


```{r}
o2groups::scale(unscaled_est[,1:3]) %>% 
  as.data.frame() %>%
  rownames_to_column(var = "group") %>%
  ggplot(aes(x = group, y = est, col = group)) +
  geom_pointrange(aes(ymin = lower_ci  , ymax = upper_ci),
                  size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed")+
  geom_point(
    data = tibble(est = o2groups::scale(delta), group = name),
    col = "black",
    shape = 18,
    size = 5
  ) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(
    title = "Estimated (colour) vs true (black) assortativity coefficients",
    x = "Group",
    y = "Estimate & 95% CI"
  )
```

## Determining the right time window
The assortativity coefficient is determined by the observed proportion of within-group transmissions and the size of each group.

As the epidemic progresses, the proportion of within-group transmissions is expected to change due to saturation (i.e. the depletion of susceptibles).
We can visualise the change in the observed proportion of within-group transmissions over time using the `binom_mix` function with different time windows.

```{r}

prop_over_time <- lapply(1:max(sim$data$date_onset), function(max_time) {
  o2groups::binom_mix(sim$data,
                      min_t = 1,
                      max_t = max_time,
                      conf_level = 0.95) %>% 
    as.data.frame() %>% rownames_to_column(var = "group") %>% 
    mutate(max_time = max_time)
}) %>% bind_rows()

prop_over_time %>% 
  drop_na() %>%
  ggplot(aes(x = max_time, y = est)) +
  geom_line(aes(col = group)) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci , fill = group), alpha = 0.2) +
  scale_color_manual(values = c("red3", "green4", "steelblue")) +
  labs(x = "Time", y = "Proportion of within-group transmissions",
       title = "Proportion of within-group transmissions over time") +
  theme_bw()

  
```

As we analyse transmission chains over time, the sample size (i.e. the number of transmission events) increases, leading to narrower confidence intervals. Additionally, the observed proportion of within-group transmissions changes due to saturation.

We can see how the estimated assortativity coefficients by estimating the assortativity coefficients over time.

```{r}
delta_est_over_time <- lapply(1:max(sim$data$date_onset), function(max_time) {
  o2groups::early_delta(sim$data,
                        min_t = 1,
                        max_t = max_time,
                        name = name,
                        size = size) %>% 
    as.data.frame() %>% rownames_to_column(var = "group") %>% 
    mutate(max_time = max_time,
           across(c(est, lower_ci, upper_ci), o2groups::scale))
}) %>% bind_rows()


delta_est_over_time %>%
  drop_na() %>%
  ggplot(aes(x = max_time, y = est)) +
  geom_line(aes(col = group)) +
  geom_ribbon(aes(ymin = lower_ci, ymax = upper_ci , fill = group), alpha = 0.2) +
  geom_hline(# true delta
    data = tibble(truth = o2groups::scale(delta), 
                  group = name),
    aes(yintercept = truth, col = group),
    lty = "dashed") +
  scale_color_manual(values = c("red3", "green4", "steelblue")) +
  labs(x = "Time", y = "Assortativity coefficient",
       title = "Estimated assortativity coefficients (ribbon) over time vs truth (dashed lines)") +
  theme_bw()
  
```

The plot above highlights how the changes in the observed proportion of within-group transmissions ( due to saturation) can lead to biased estimates of the true assortativity coefficients (dashed lines).
