---
output: github_document
---

# Overview

[Assortativity](https://en.wikipedia.org/wiki/Assortativity) refers to the tendency of individuals to interact preferentially with others who are similar to them in certain attributes. The assortativity coefficient `delta` plays a crucial role in modeling epidemics by capturing the mixing patterns within a population. In the context of epidemics, assortativity specifically measures the extent to which infections are more likely to spread within certain groups compared to others. A `delta` value greater than 1 indicates that individuals are more likely to interact and transmit infections within their own group, while a `delta` value below 1 suggests a tendency for infections to spread across different groups.

The assortativity coefficient has significant implications for epidemic dynamics. High assortativity can lead to concentrated outbreaks within specific groups, potentially resulting in faster spread and larger clusters of infected individuals. Conversely, low assortativity promotes more even mixing between groups, which may lead to a more widespread but potentially slower epidemic.

Changes in assortativity over time can indicate shifts in behavior or social interactions that may impact the spread of the disease. As a result, the assortativity coefficient can serve as an informative indicator of the effectiveness of non-pharmaceutical interventions, such as physical distancing measures or targeted interventions within specific groups.

The `o2groups` package provides a framework for modeling outbreaks using branching processes specifically designed for scenarios involving multiple groups.

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

# Installation

```{r}
#devtools::install_github("CyGei/o2groups")
library(o2groups)
#library(tidyverse)
```

# Example

## Simulation

We first specify the input parameters for our simulation.

```{r}
n_simulations = 100
n_cores = parallel::detectCores() - 2
duration = 100
n_groups = 4
size = c(1e3, 1e4, 1e3, 1e3)
name = LETTERS[1:n_groups]
delta = c(10, 2, 1, 4)
intro_group = LETTERS[1:n_groups]
intro_n = rep(10, n_groups)
r0 = c(3, 4, 2, 3.5)
generation_time = simulacr::make_disc_gamma(mean = 5, sd = 2)$d(1:30)
```

We then run the outbreak simulation 100 times to account for stochasticity.

```{r}
set.seed(123)

  out <-
  simulate_groups_furrr(
    n_simulations = n_simulations,
    duration = duration,
    n_groups = n_groups,
    size = size,
    name = name,
    delta = delta,
    intro_group = intro_group,
    intro_n = intro_n,
    r0 = r0,
    generation_time = simulacr::make_disc_gamma(mean = 5, sd = 2)$d(1:30) )

data = out$data
stats = out$stats
```

## Depletion of susceptibles

```{r}
pal = c("A" = "magenta", "B" = "orange", "C" = "forestgreen", "D" = "steelblue")
gg_col <- list(
  scale_color_manual(values = pal),
  scale_fill_manual(values = pal)
)
gg_scale <- list(scale_x_continuous(breaks = seq(0,100,10)))


#plot_tree(subset(data, simulation == 1), pal = pal)
#plot_stats(stats)[[1]] + gg_col + gg_scale
plot_stats(stats)[[2]] + gg_col  + gg_scale

```

## Case Reproduction Number

```{r message=FALSE}
# Indiv Ri
Ri <-
  map(.x = unique(data$simulation), ~ get_Ri(data[data$simulation == .x,])) %>%
  bind_rows(., .id = "simulation")

Ri_long <- Ri %>%
  pivot_longer(cols = LETTERS[1:4],
               names_to = "target") %>%
  select(simulation, id, group, target, value, date_infection)

# Summary Ri
Ri_sims <- Ri_long %>%
  group_by(group, target, date_infection, simulation) %>%
  summarise(Ri = mean(value))
Ri_means <- Ri_sims %>%
  group_by(group, target, date_infection) %>%
  summarise(Ri = mean(Ri))

# Total Ri
total_Ri_sims <- Ri %>%
  group_by(simulation, group, date_infection) %>%
  summarise(Ri = mean(Ri))

total_Ri_means <- Ri %>%
  group_by(group, date_infection) %>%
  summarise(Ri = mean(Ri))
```

```{r}

# plot
p_facet_Ri <- ggplot(data = Ri_sims,
                     aes(x = date_infection,
                         y = Ri,
                         col = group)) +
  facet_grid(target ~ group, scales = "free_y") +
  geom_point(alpha = 0.5) +
  geom_line(data = Ri_means,
            col = "black") +
  theme_bw()+
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.title.x = element_blank(),
        legend.position = "none") +
  gg_col +
  gg_scale+
  scale_y_continuous(breaks = 0:10)


p_total_Ri <- ggplot(total_Ri_sims,
                     aes(x = date_infection,
                         y = Ri,
                         col = group,
                         group = group)) +
  geom_point() +
  geom_line(data = total_Ri_means,
            col = "black") +
  geom_hline(data = tibble(group = LETTERS[1:4],
                           r0 = r0 ),
             aes(yintercept = r0, group = group),
             col = "#4C4E52", lty = "solid")+
  geom_hline(aes(yintercept = 1), col = "#4C4E52", lty = "dotted")+
  facet_grid(~ group) +
  theme_bw()+
  theme( strip.text.x = element_blank(),
         legend.position = "none")+
  gg_col +
  gg_scale+
  scale_y_continuous(breaks = 0:10)

patchwork::wrap_plots(p_facet_Ri,p_total_Ri, ncol = 1, guides = "collect",
                      heights = c(3,1))
```

The plot above describes the average number of secondary cases a primary case from a given group (columns) generates in every group (rows) over time.

## Mixing 

```{r, message=FALSE}
# compute the mixing proportions at every timepoint
mix_t <- map(.x = unique(data$simulation),
             function(sim) {
               sim_data <- filter(data, simulation == sim)
               map(.x = unique(sim_data$date_infection),
                   function(date) {
                     mix <- sim_data %>% get_mixing(., min_t = date, max_t = date)
                     return(mix$result)
                   }) %>%
                 bind_rows(., .id = "t") %>%
                 mutate(t = as.integer(t))
             }) %>%
  bind_rows(., .id = "simulation") %>%
  mutate(across(c(source_group, group),
                ~ factor(., levels = LETTERS[1:4])))

mean_freqs <- mix_t %>%
  group_by(t, source_group, group) %>%
  summarise(freq = mean(freq))

# inital mixing frequencies
Truth <- generate_Mcol(
  n_groups = n_groups,
  size = size,
  name = LETTERS[1:4],
  delta = delta
) %>%
  as.data.frame(.) %>%
  rownames_to_column(var = "group") %>%
  pivot_longer(-group, names_to = "source_group", values_to = "value") %>%
  select(source_group, group, value) %>%
  arrange(source_group)


ggplot(mix_t) +
  aes(x = t,
      y = freq ,
      col = source_group) +
  facet_grid(group ~ source_group) +
  geom_point() +
  geom_line(data = mean_freqs, col = "#4C4E52") +
  geom_hline(data = Truth,
             aes(yintercept = value), lty = "dotted") +
  theme_bw() +
  theme(legend.position = "none") +
  gg_col +
  gg_scale

```

The plot above illustrates the temporal evolution of group infection mixing. Each data point corresponds to the percentage of infections originating from a particular group (column) and spreading to another group (row) at time t. The solid grey line represents the average infection mixing frequency, while the dotted line represents the initial mixing frequency.

## Retrieve `delta`

The previous plot highlighted how saturation affects the infection mixing proportions.

### Early Phase

`Delta` values can be derived from the groups' sizes and the observed mixing frequencies during the initial stage of the epidemic, when most of the groups' individuals are susceptible.

Note that it is important to find the right early phase window to retrieve the appropriate mixing matrix (`Mcol_observed`).

```{r}
#early phase
Mcol_observed <- get_mixing(data, min_t = 5, max_t = 12)$Mcol

Mcol_to_delta(n_groups = n_groups,
              size = size,
              name = LETTERS[1:4],
              Mcol_observed = Mcol_observed) %>% round()


```

### At any time point

In addition, we can estimate delta at any given time point by incorporating additional information such as the group's basic reproduction number (R0) and the proportion of susceptible individuals at a specific time (t). The code below estimates delta over every time point.

```{r, warning=FALSE, message=FALSE}
timepoints <- 1:100
output <- matrix(NA, nrow = length(timepoints), ncol = n_groups)

for (tp in timepoints) {
  Mcolt_observed <- get_mixing(data, min_t = tp, max_t = tp)$Mcol

  prop_sus <- stats %>%
    filter(time == tp) %>%
    group_by(group) %>%
    summarise(prop_susceptible = mean(prop_susceptible)) %>%
    pull(prop_susceptible)

  delta_optim <- Mcolt_to_delta_optim(
    n_groups = 4,
    size = size,
    name =name,
    r0 = r0,
    prop_susceptible = prop_sus,
    Mcolt_observed = Mcolt_observed
  )

  output[tp, ] <- delta_optim

}
colnames(output) <- name

```

However, it is important to note that estimating `delta` during the very early or late stages of the epidemic may be less reliable due to potential limitations in sample size.

```{r}
output %>%
  as_tibble() %>%
  mutate(timepoint = row_number()) %>%
  filter(timepoint <= 40) %>%
  pivot_longer(cols = -timepoint, names_to = "group", values_to = "delta") %>%
  ggplot(aes(x = timepoint, y = delta, col = group))+
  geom_point()+
  geom_line()+
  geom_hline(data = tibble(delta = delta), aes(yintercept = delta))+
  gg_col+
  theme_bw()
```
